---
title: "Workshop 8 TEMPLATE"
subtitle: "9:00 am"
author: "Sascha Roberts"
date: "2026-05-22"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 7 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest"
)


```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
    "Lotus scoparius" = lotsco_col,
    "Eschscholzia californica" = esccal_col,
    "Grindelia camporum" = gricam_col,
    "Stipa pulchra" = stipul_col,
    "Penstemon centranthifolius" = pencen_col,
    "Salvia leucophylla" = salleu_col,
    "Encelia californica" = enccal_col
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# create new object for clean data
drought_exp_clean <- drought_exp |> 
  # taking out spaces and capitals in column names
  clean_names() |> 
  # creating new column to store binomial species name
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # creating new column with full name for treatment
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # reordering species levels
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  # reordering water treatment level
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting desired columns 
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: checking_structure

str(drought_exp_clean)

```



Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: 10_rows

# displaying 10 rows from data frame
slice_sample (drought_exp_clean, n = 10)

```




# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis. 

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 


```{r}
#| label: scatter_plot

#baselayer: ggplot
ggplot(data = drought_exp_clean,
       # x-axis: sla
       # y-axis: total plant mass
       mapping = aes(x = sla,
                     y = total_g)) +
  # first layer: geompoint
  geom_point () + 
  # renaming axis and title
  labs (x = "Specific Leaf Area (cm\U00B2/g)",
        y = "Total Plant Mass (g)",
        title = "As Specific Leaf Area increases, Total Mass increases")

```



## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.


```{r}
#| label: jitterplot

# baselayer: ggplot
ggplot(data = drought_exp_clean,
       # x-axis: water treatment
       # y-axis: total plant mass
       # color by treatment
       mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment)) +
  # first layer: geom jitter
  # width is 0.1, no point movement on y-axis
  geom_jitter(height = 0,
              width = 0.1) + 
  # setting custom color for water treatment
  scale_color_manual (
    values = c("Drought stressed" = ds_col,
               "Well watered" = ww_col)) +
  # renaming axis and title
  labs(x = "Water Treatment",
       y = "Total Mass (g)",
       title = "Median plant mass of Well Watered is higher \nthan Drought Stressed") + 
  # removing legend
  theme (legend.position = "none")

```




## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass.


```{r}
#| label: jitterplot_2

# baselayer: ggplot
ggplot(data = drought_exp_clean,
       # x-axis: species
       # y-axis: total plant mass
       # color by species
       mapping = aes(x = species_name,
                     y = total_g,
                     color = species_name)) +
  # first layer: geom jitter
  # width is 0.1, no point movement on y-axis
  geom_jitter(height = 0,
              width = 0.1) + 
  # setting custom color for species
  scale_color_manual (values = c(
    "Lotus scoparius" = lotsco_col,
    "Penstemon centranthifolius" = pencen_col,
    "Salvia leucophylla" = salleu_col,
    "Encelia californica" = enccal_col,
    "Stipa pulchra" = stipul_col,
    "Eschscholzia californica" = esccal_col,
    "Grindelia camporum" = gricam_col
  )) +
  # wrap species labels on x-axis
  scale_x_discrete(labels = label_wrap(10)) +
  # renaming the axis and title
  labs(x = "Species Name",
       y = "Total Mass (g)",
       title = "Comparing Plant Mass of Species") + 
  theme (legend.position = "none")

```




# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1, # fromula: no predictors
  data = drought_exp_clean # data = clean
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name,
  data = drought_exp_clean
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean
)

# model 5: SLA
model5 <- lm(
  total_g ~ sla,
  data = drought_exp_clean
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species
model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)

```




## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  
```{r}
#| label: calculating_aic

# calculating AIC
AICc (
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  # arranging output in order of AIC
  arrange (AICc)

```

Arrange by decreasing AIC.




Insert code to display the summary for the best model. 

```{r}
#| label: model_summary

summary (model4)

```

- (Intercept): represents model prediction for plant mass of lotus scoparius in the drought stressed treatment
- Water_treatmentWell watered: model prediction for the difference in Lotus scoparius mass between the well watered and drought stressed treatment
- species_namePenstemon centranthifolius: model prediction for the difference between Penstomn mass in the well watered treatment and Lotus mass in the well watered treatment.
- species_nameEschscholzia californica: model prediction for the difference between Eschscholzia mass in the well watered treatment and Lotus mass in the well watered treatment. 

## Model diagnostics

```{r}
#| label: check_diagnostics

#displaying plots in 2 by 2 grid
par(mfrow = c(2, 2))
# generating diagnostic plots
plot(model4)

```

 - residuals appear normally distributed
 - residuals appear homoscedastic (have constant variance)
 - no clear outliers in data set influencing model coefficient estimates



## Stop and think: what does this model mean?  

**What is the best model?**

The best model include watewr treatment and species as predcitors.

**How much variation in the response (total mass, in grams) does this model explain?**

72.6% of the variation in plant mass is explained by a model with both of these predcitors.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- Treatment influences plant mass.
- Species influences plant mass.

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}

preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  rename(water_treatment = x,
         species_name = group)

```



# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}

ggplot( data = drought_exp_clean,
        mapping = aes (x = water_treatment,
                       y = total_g,
                       color = water_treatment)) +
  geom_jitter(height = 0,
              width =0.1,
              shape = 21,
              alpha = 0.8) +
  geom_pointrange (data = preds,
                   mapping = aes(x = water_treatment,
                                 y = predicted,
                                 ymin = conf.low,
                                 ymax = conf.high)) +
  facet_wrap (~ species_name) + 
  theme (legend.position = "none")

```

# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







